V4 Neural Network Model for Shape-based Feature Extraction and Object Discrimination

### 2015-09-02 article.tex
brief sec 1 and sec2, deep learning added

Primate brains possess two distinct visual pathways \cite{ettlinger1990,lehky2007}.
The dorsal pathway is involved with processing the object's spatial location. 
The ventral pathway is involved with object recognition.
In this paper, we concentrate on the latter, the ventral pathway, and visual area V4 in particular.
V4 lies in the middle of the ventral pathway (\figurename~\ref{fig:1}).
It extracts the feature of contour conformation, which represents a middle level abstraction of visual information in the brain.
Lower levels (V1 and V2) extract features such as edges and orientations \cite{hubel1962,hubel1965}.
Higher levels (inferior temporal cortex) comprehend concepts of complex objects like faces and body parts \cite{bruce1981,bell2009}.
As an intermediate stage, V4 plays a crucial role in transforming low-level orientation signals into complex object representations.

@@

In this paper, we propose a multi-layer neural network model which imitates the neural mechanism of visual area V4 in the ventral visual pathway.
Different stages of the neural pathway in the visual cortex are explicitly mapped to different layers of the neural network model.
In this feed-forward network, features are extracted hierarchically from input images. 
High-level local image features emerge in the output layer of the network. 
The features encode local structures of object shapes in a way which closely resembles the mechanism of the cortex 
and therefore achieve a good balance between the discrimination and generalization abilities.
Moreover, this multilayer neural network shows a successful attempt to integrate different stages of visual processing in the brain. 
It provides an extensible framework so that future findings in cortex mechanisms can be adopted and applied to computer vision tasks.

We briefly introduce some related work on image features and V4 models in section~\ref{sec:2}.
In section~\ref{sec:3}, we give the detailed description of our proposed model.
It is capable of locating salient points and generating discriminative local shape features around salient points.
In section~\ref{sec:4}, we show several experiments in which we evaluate the proposed model.
The model is compared with real V4 neurons in a shape distinction experiment.
The features extracted by the model are applied to various computer vision tasks.
The conclusion is summarized in section~\ref{sec:5}.

@@

In this section, we briefly explain the difference between our model and traditional image features.
We also review several existing models of V4 and some other related models.

\subsection{Previous V4 models}

V4 neurons have been examine with different varieties of visual stimuli, including light bars, sinusoidal gratings~(\figurename~\ref{fig:2a}),
non-Cartesian gratings~(\figurename~\ref{fig:2b}), closed shapes~(\figurename~\ref{fig:2c}) \cite{desimone1987,gallant1996,pasupathy1999,pasupathy2001}.
They generally prefer complex features to simple features such as bars and sinusoidal gratings \cite{gallant1996}.
Most V4 neurons are tuned for the curvature and orientation of boundary fragments \cite{pasupathy2001}.
Therefore, unlike neurons in V1, V4 neurons can not be modeled with linear filters.

@@

Several models have been proposed for V4 neurons.
We briefly introduce the SRF model \cite{david2006} and the HMAX model \cite{riesenhuber1999,cadieu2007}.
The spectral receptive field (SRF) \cite{david2006} describes properties of V4 receptive field in terms of the orientation and spatial frequency spectrum.
This model is powerful in describing the shape selectivity of V4 neurons.
However, it does not provide any computational or neural mechanism to achieve V4 selectivity via spectral receptive field.
The HMAX model was first proposed in \cite{riesenhuber1999} as a generic model for object recognition in the visual cortex.
It models the visual cortex as a hierarchical architecture consisting of cascaded linear filters and non-linear maximum pooling operations.
It has been also adopted as a model for V4 shape selectivity and invariance \cite{cadieu2007}.
The training of the model is an NP-complete problem. 
The authors used a greedy algorithm to obtain approximated solutions but they did not provide any biological evidence for the algorithm.

@@

According to the hierarchy of the ventral visual pathway,
area V4 receives input from the lower levels including area V1 and V2.
These areas have been well studied since 1960s 
by Hubel, Wiesel \cite{hubel1962,hubel1965} and succeeding researchers.

@@

>>>>

Primate brains possess two distinct visual pathways \cite{ettlinger1990}.
The dorsal pathway is involved with processing the object's spatial location. 
The ventral pathway is involved with object recognition.
In this paper, we concentrate on the latter, the ventral pathway, and visual area V4 in particular.
V4 lies in the middle of the ventral pathway (\figurename~\ref{fig:1}).
It extracts the feature of contour conformation, which represents a middle level abstraction of visual information in the brain.
Lower levels (V1 and V2) extract features such as edges and orientations \cite{hubel1962}.
Higher levels (inferior temporal cortex) comprehend concepts of complex objects like faces and body parts \cite{bell2009}.
As an intermediate stage, V4 plays a crucial role in transforming low-level orientation signals into complex object representations.

@@

In this paper, we propose a multi-layer neural network model which imitates the neural mechanism of visual area V4 in the ventral visual pathway.
Different stages of the neural pathway in the visual cortex are explicitly mapped to different layers of the neural network model.
In this feed-forward network, features are extracted hierarchically from input images.
High-level local image features emerge in the output layer of the network. 
The features encode local structures of object shapes in a way which closely resembles the mechanism of the cortex.
With a limited number of training images, the model can learn to distinguish between different categories of objects.
Moreover, this multilayer neural network shows a successful attempt to integrate different stages of visual processing in the brain. 
It provides an extensible framework so that future findings in cortex mechanisms can be adopted and applied to computer vision tasks.

We briefly introduce some related work on image features and V4 models in section~\ref{sec:2}.
In section~\ref{sec:3}, we give the detailed description of our proposed model.
It is capable of locating salient points and generating discriminative local shape features around salient points.
In section~\ref{sec:4}, we show several experiments in which we evaluate the proposed model.
The model is compared with real V4 neurons in a shape distinction experiment.
The features extracted by the model are applied to various computer vision tasks.
The discussion and conclusion is summarized in section~\ref{sec:5}.

@@

In this section, we briefly review some existing models of V4 and other related models.

V4 neurons have been examine with different varieties of visual stimuli, including light bars, sinusoidal gratings~(\figurename~\ref{fig:2a}),
non-Cartesian gratings~(\figurename~\ref{fig:2b}), and closed shapes~(\figurename~\ref{fig:2c}) \cite{gallant1996,pasupathy2001}.
They generally prefer complex features to simple features such as bars and sinusoidal gratings \cite{gallant1996}.
Most V4 neurons are tuned for the curvature and orientation of boundary fragments \cite{pasupathy2001}.
Therefore, unlike neurons in V1, V4 neurons can not be modeled with linear filters.

@@

Several models have been proposed for V4 neurons.
We briefly introduce the SRF model \cite{david2006} and the HMAX model \cite{cadieu2007}.
The spectral receptive field (SRF) \cite{david2006} describes properties of V4 receptive field in terms of the orientation and spatial frequency spectrum.
This model is powerful in describing the shape selectivity of V4 neurons.
However, it does not provide any computational or neural mechanism to achieve V4 selectivity via spectral receptive field.
The HMAX model was first proposed in \cite{riesenhuber1999} as a generic model for object recognition in the visual cortex.
It models the visual cortex as a hierarchical architecture consisting of cascaded linear filters and non-linear maximum pooling operations.
It has been also adopted as a model for V4 shape selectivity and invariance \cite{cadieu2007}.
Training this model is an NP-complete problem. 
The authors used a greedy algorithm to obtain approximated solutions but they did not provide any biological evidence for the algorithm.

The deep learning model also draws inspiration from the layered structure of the cortex \cite{bengio2009}.
It learns representations of data through stacked layers of Restricted Boltzmann Machines (RBM).
When it is trained with natural images, 
the kernels of the data-connected layer resemble the pattern of the receptive field of neurons in the primary visual cortex.
By training many stacked layers of RBM, highly abstracted features can be extracted automatically from input images.
Therefore, with additional fine tuning, the deep model exhibits high performance in image classification \cite{krizhevsky2012}.
Our proposed model employs a similar multilayer structure but combines automatic feature extraction with artificial features.
This combined strategy allows our model to work on fewer training images.

@@

According to the hierarchy of the ventral visual pathway,
area V4 receives input from the lower levels including area V1 and V2.
These areas have been well studied since 1960s 
by Hubel, Wiesel \cite{hubel1962} and succeeding researchers.

@@

This process requires a large number of training samples (\cite{krizhevsky2012} uses 1.2 million training images)
and a neural network of very large scale (In \cite{krizhevsky2012}, the network has 8 layers and more than 400,000 neurons).
Deep learning can be viewed as automatic feature extraction.
In our proposed model, we combine automatic feature extraction with artificial feature extraction.
With prior knowledge of the cortical neurons, we manually design the layers corresponding to the primary visual cortex
and V4 features are automatically extracted based on the output of the artificially designed network.
This strategy saves us from retraining the first layers every time
and it allows the model to work on limited number of training images.
